Ornaments as indicators of social changes in northeastern Taiwan before and after European colonial period

### 2019-12-14 analysis/paper/paper.Rmd
change title, fix typo, and change spelling

---
title: "Ornaments as indicators of social changes and cultural practices in northeastern Taiwan before and after European colonial period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      #reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, paleolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages [@Liu2011; @Chen2017]. It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeast Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later the global trade network with the Europeans in the 17th century brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected to trade networks via visits of other indigenous groups, Chinese merchants, and Europeans, via sea.

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Keelung in 1626, and Fort San Domingo in 1629 at Tamsui. They sent missionaries to local indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of indigenous communities. A Dominican priest in 1632 reported that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compensation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an indigenous funeral document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tributes [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotations with indigenous communities to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the exotic materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, an academic field survey for plains indigenous groups reported that fish-shaped metal necklaces necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. Although these historical records are fragmentary and may contain some biases [@Galloway2006] that have not yet been studied in detail, we find consistency among multiple sources in their descriptions of how ornaments represent high status or specialised social roles in indigenous communities in Yilan. Compared to the European period, there are fewer documentary mentions of beads in the Chinese period and the descriptions are limited to clothing, but these generally confirm the role of beads as status markers.

@@

Information about Kiwulan (Figure \@ref(fig:KWL-map)) comes from a rescue archaeology project that during 2001-2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artefacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-term settlement. Artefact locations were recorded to the 2 x 2 m sub-square they were recovered in; they lack individual point provenance. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of A and D sections, which is also the study area where our samples selected from. In the AD area, post-holes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction (Figure \@ref(fig:KWL-sam-area)).

The chronology of Kiwulan can be divided into two phases represented by a upper component (1400-1900 AD, 600-100 BP) and a lower component (700-1200 AD, 1200-800 BP) separated by a sterile deposit spanning c. 150 years. These component divisions are based on the differences in the colour and texture of the deposit. The interpration of the sterile deposit is still under debate, with pollen analysis suggesting dry weather leading to site abandonment [@Lin2015; @Chiu2004; @Chen2007]. There are 32 radiocarbon ages spanning the two components, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). We focus on the upper component because only this componant spans the periods of pre-European contact, European presence, and Chinese presence. In the upper component, all excavation squares in our sampling area show signs of continuous human occupation during each of the three phases. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the radiocarbon dates, excavation depth, consistency of contexts, and types of chronologically diagnostic ceramics such as blue and white porcelains [@Hsieh2009; @Wang2011]. However, we found some ambiguities in the previous chronology, so to help answer our research question, we re-examined the upper component to devise a new chronology to assign artefacts to the pre-European, European, and the Chinese periods.

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan. Frequency represents artefact counts", fig.width=image_width_1_col_inch}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories 
ornaments_count <-
  ornaments_period %>% 
  filter(Categories != "Unknown Metal") %>%
  ungroup() %>% # need to upgroup because of rowwise or factor in previous chunk?
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  group_by(Categories) %>%
  count() %>% 
  ungroup() %>% 
  mutate(Categories = fct_reorder(Categories, n))
plot_ornaments_count <- 
ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 160)) +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 10)
ggsave(plot = plot_ornaments_count,
       here("analysis", "figures",
       "plot-ornaments-count.jpg"),
       width = 90, 
       height = 40,
       dpi = 300,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count.jpg"))
```

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "small bead", "small (0.5-1 cm)", Type),
         Type = ifelse(Type == "glass bead", "medium (1-2 cm)", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artefact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)

@@

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualisations contained in this paper is included in http://doi.org/10.17605/OSF.IO/R8YGA. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualisations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequencies of the major classes of ornaments for different time periods at Kiwulan. The difference in frequencies between the three time periods reflect significant differences in the use of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$). Most ornament types were present before European contact. Ornament frequencies reached a peak during the European period and then dropped during the Chinese period, especially gold-foil beads. This trend can be also seen on other ornaments including carnelian beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period. To model the number of ornaments as a function of the mass of ceramics in each period, a Poisson GLM with a log link function was used. The model reveals that ceramic abundance strongly predicts the number of ornaments (β = `r beta_print`, p = `r pval_print`). If ceramic abundance is a suitable proxy for population at Kiuwan due to its basic role as cooking vessels, then ornament quantities per period may be influenced by the number of people living at the site.

@@

Point pattern analysis can assess whether the distribution of artefacts represents hotspots produced by non-random processes [@Bevan2016; @Ducke2015], such as concentrations of ornaments in specific households that might result from social inequality stimulated by a colonial presence. To prepare the ornament location data for point pattern analysis, because artefacts from Kiwulan lack exact piece-provance data, we assigned each ornament to a random coordinate pair in the square it was recovered from. The next step was to divide the ornaments into three time periods. Finally we computed the kernel densities for each time period for comparison. Kernel density estimations (KDE) compute the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here we use KDE to visualize core areas of ornaments and surrounding neighbourhoods [@Bonnier2019; @Cortegoso2016]. Density values of artefacts per square meter were calculated for each cell.

@@

To test for randomness in spatial locations, we used a Monte Carlo method to simulate average nearest-neighbour distances (ANN). Figure \@ref(fig:hypo-test-distribution) shows the observed ANN distances with the distributions of the ANN distances calculated on 1000 simulations of random ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European period, which means the ornaments have non-randomly clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese period, with about one third of the simulated values are greater than our observed ANN value. The Chinese period has fewer artefacts in any category, likely reflecting a smaller population at Kiwulan at this time, making spatial patterns and hotspots difficult to discern with confidence. This testing reveals clustering of ornaments during the European period is highly non-random, potentially indicating different degrees of access to foreign ornaments or a concentration of power to control the distribution of ornaments at Kiwulan during this period.

@@

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. The corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999]. We may be able to interprete Yilan social organisation as moving from corporate mode before the European arrival, to a network mode during European presence. The small number of ornaments, and less concentrated distribution during the pre-European period appears consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode, and the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

@@

We are still far from understanding the full variety of colonial impacts on peripheral indigenous communities. The origin of ornaments and how their sources changed over time would give more information to construct a clear picture of complex trade networks during this periods. By focusing on the distribution patterns in a settlement site, the Kiwulan ornaments suggest that foreign ornaments can be a proxy to detect indirect colonial influence on local indigenous populations. Ornaments give insights into the amplification of social inequality stimulated by European colonisation. It also shows the agency of indigenous people to incorporate ornaments into their social system and use them in their daily lives to display or intensify status differences. Future work could extend this approach to sourcing the ornaments directly with geochemical methods, and studies of other trade goods such as ceramics. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and future tests of this should include analysis of pottery production and standardisation, and mortuary practices.

>>>>

---
title: "Ornaments as indicators of social changes in northeastern Taiwan before and after European colonial period"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      #reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonization, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonization; Trade ornaments; Exchange networks; Taiwan; spatial pattern
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

The island of Taiwan lies at the junction of mainland China, Southeast Asia, and Northeast Asia in the Pacific Ocean. The prehistory of Taiwan island could be roughly divided into three major periods, palaeolithic (c. 27,000 BP- 5000BP), Neolithic (c. 6500- 2000BP), and Iron age (c. 2000- 400BP) with slightly regional differences in onset of each period and variations in style of artifacts and assemblages [@Liu2011; @Chen2017]. It is generally accepted that Taiwan entered the historical period in the early 17th century due to the colonial activities of the Spanish and the Dutch who played an important role in keeping written records about Taiwan. The European colonial presence in Taiwan ended in 1662 when the Dutch were defeated by the kingdom of Tungning, founded by Koxinga from China. Later in 1683, Taiwan was incorporated into the Qing dynasty in China and a large wave of Han Chinese migrated to Taiwan during the late 18th century. Because of natural safe harbors, northeast Taiwan was involved in a regional trade network through cross-culture interactions with Chinese merchants since the 14th century, and later the global trade network with the Europeans in the 17th century brought more trade goods circulated in Southeast Asia into Taiwan [@Chen2005; @Liu2017]. Although located on the periphery of regional trade centers, Yilan was connected to trade networks via visits of other indigenous groups, Chinese merchants, and Europeans, via sea.

The European presence in northern Taiwan started with the Spanish who founded Fort San Salvador at Keelung in 1626, and Fort San Domingo in 1629 at Tamsui. They sent missionaries to local indigenous settlements in this region [@Blusse2000, pp. 343] and kept records about their observations of indigenous communities. A Dominican priest in 1632 reported that the Taparri, an indigenous tribe from northern Taiwan, exchanged carnelian beads with other indigenous groups. This form of exchange was widespread and even the Spanish soldiers used carnelian beads as bargaining chips for gambling [@LiandWu2006, pp. 132-149]. The use of beads as prestige goods is further indicated by their role in bride price payments, and compenzation to resolve disputes [@LiandWu2006, pp. 132-149]. Other records mention that the female shamans in the tribe would use carnelian beads as magical items in ritual healing practices [@Borao2009, pp. 122-151]. Records of an indigenous funeral document the use of carnelian beads in ritual contexts, with more carnelian beads, pottery, and cloth placed into the graves of more influential people to indicate their family's higher status [@LiandWu2006, pp. 153]. While a full critical analysis of these historical accounts remains to be produced, we take them to minimally indicate that carnelian beads were already treated as prestige goods in Yilan before the arrival of Europeans. In 1642, the Dutch Vereenigde Oostindische Compagnie (VOC) defeated the Spanish and took over their forts in northern Taiwan. They introduced a feudal system in an attempt to control the indigenous communities by asking indigenous leaders to attend an annual ceremony for demonstrating their loyalty and paying tributes [@Andrade2007, ch. 9; @Kang2016, ch. 4]. The Dutch provided beads and other goods based on negotations with indigenous communities to secure alliances in the annual ceremony or during travelling [@Kang2016, ch. 6]. We might predict that the activities of the Dutch feudal system to build and maintain alliances resulted in an increase in the amount and diversity of ornaments in northeastern indigenous communities during this period.

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty (1616-1911) contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, fish-shaped necklaces made of metal threads had high value due to their delicacy and the exotic materials invested in production. These were usually possessed by wealthy people. Other people wore carnelian beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, an academic field survey for plains indigenous groups reported that fish-shaped metal necklaces necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232]. Although these historical records are fragmentary and may contain some biases [@Galloway2006] that have not yet been studied in detail, we find consistency among multiple sources in their descriptions of how ornaments represent high status or specialized social roles in indigenous communities in Yilan. Compared to the European period, there are fewer documentary mentions of beads in the Chinese period and the descriptions are limited to clothing, but these generally confirm the role of beads as status markers.

@@

Information about Kiwulan (Figure \@ref(fig:KWL-map)) comes from a rescue archaeology project that during 2001-2004 in advance of a water diversion project and road bridge construction. The excavations used 2 mm and 1.5 mm mesh screens and covered eight open area sections in total of 262 squares (4 m by 4 m) reaching 3,814 m^2^ [@Chen2007]. The nearly 2 m thick archaeological deposits reveal a large amount of artifacts, burials, middens, post-holes, wooden pillars, and stone structures, all of which indicates it was a long-term settlement. Artifact locations were recorded to the 2 x 2 m sub-square they were recovered in; they lack individual point provenance. Based on the continuity of deposition and the frequency of artifacts, the center of the site is the open area consisting of A and D sections, which is also the study area where our samples selected from. In the AD area, post-holes were found aligned in a north-south direction in some intervals with construction marks, which were interpreted as the remains of silt house structures. At the north margin of the dwelling place were burials that are mostly oriented in an east-west direction (Figure \@ref(fig:KWL-sam-area)).

The chronology of Kiwulan can be divided into two phases represented by a upper component (1400-1900 AD, 600-100 BP) and a lower component (700-1200 AD, 1200-800 BP) separated by a sterile deposit spanning c. 150 years. These component divisions are based on the differences in the colour and texture of the deposit. The interpretation of the sterile deposit is still under debate, with pollen analysis suggesting dry weather leading to site abandonment [@Lin2015; @Chiu2004; @Chen2007]. There are 32 radiocarbon ages spanning the two components, previously published by @Chen2007, and shown here in Figure \@ref(fig:C14-data) and Table \@ref(tab:C14-table-data). We focus on the upper component because only this component spans the periods of pre-European contact, European presence, and Chinese presence. In the upper component, all excavation squares in our sampling area show signs of continuous human occupation during each of the three phases. Previous work divided the upper component into six analytical units, spanning from the 14th century to the 19th century, according to the radiocarbon dates, excavation depth, consistency of contexts, and types of chronologically diagnostic ceramics such as blue and white porcelains [@Hsieh2009; @Wang2011]. However, we found some ambiguities in the previous chronology, so to help answer our research question, we re-examined the upper component to devise a new chronology to assign artifacts to the pre-European, European, and the Chinese periods.

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan. Frequency represents artifact counts", fig.width=image_width_1_col_inch}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories 
ornaments_count <-
  ornaments_period %>% 
  filter(Categories != "Unknown Metal") %>%
  ungroup() %>% # need to upgroup because of rowwise or factor in previous chunk?
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  group_by(Categories) %>%
  count() %>% 
  ungroup() %>% 
  mutate(Categories = fct_reorder(Categories, n))
plot_ornaments_count <- 
ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 160)) +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 10)
ggsave(plot = plot_ornaments_count,
       here("analysis", "figures",
       "plot-ornaments-count.jpg"),
       width = 90, 
       height = 40,
       dpi = 300,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count.jpg"))
```

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "small bead", "small (0.5-1 cm)", Type),
         Type = ifelse(Type == "glass bead", "medium (1-2 cm)", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artifact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)

@@

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in http://doi.org/10.17605/OSF.IO/R8YGA. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

Figure \@ref(fig:plot-counts-by-period) shows the comparison of frequencies of the major classes of ornaments for different time periods at Kiwulan. The difference in frequencies between the three time periods reflect significant differences in the use of ornaments (chi-square = `r chisq_test_result_stat`, df = `r chisq_test_result_df`, p-value = $`r chisq_test_result_p_value`$). Most ornament types were present before European contact. Ornament frequencies reached a peak during the European period and then dropped during the Chinese period, especially gold-foil beads. This trend can be also seen on other ornaments including carnelian beads, metal rings, and bells. However, glass beads show a different pattern that indicates a higher frequency in the pre-European contact, and then a decrease in the European period and a further decrease in the Chinese period. To model the number of ornaments as a function of the mass of ceramics in each period, a Poisson GLM with a log link function was used. The model reveals that ceramic abundance strongly predicts the number of ornaments (β = $`r beta_print`$, p = $`r pval_print`$). If ceramic abundance is a suitable proxy for population at Kiuwan due to its basic role as cooking vessels, then ornament quantities per period may be influenced by the number of people living at the site.

@@

Point pattern analysis can assess whether the distribution of artifacts represents hotspots produced by non-random processes [@Bevan2016; @Ducke2015], such as concentrations of ornaments in specific households that might result from social inequality stimulated by a colonial presence. To prepare the ornament location data for point pattern analysis, because artifacts from Kiwulan lack exact piece-provance data, we assigned each ornament to a random coordinate pair in the square it was recovered from. The next step was to divide the ornaments into three time periods. Finally we computed the kernel densities for each time period for comparison. Kernel density estimations (KDE) compute the probability of the density of ornaments across space by creating a continuous, smooth density surface across space. Here we use KDE to visualize core areas of ornaments and surrounding neighbourhoods [@Bonnier2019; @Cortegoso2016]. Density values of artifacts per square meter were calculated for each cell.

@@

To test for randomness in spatial locations, we used a Monte Carlo method to simulate average nearest-neighbour distances (ANN). Figure \@ref(fig:hypo-test-distribution) shows the observed ANN distances with the distributions of the ANN distances calculated on 1000 simulations of random ornament locations. The results show that 100% of the simulated values are much greater than our observed ANN value during the European period, which means the ornaments have non-randomly clustered distributions. A similar, but less extreme, result is also observed during the pre-European period. The observed distribution of ornaments is more similar to the random distributions during the Chinese period, with about one third of the simulated values are greater than our observed ANN value. The Chinese period has fewer artifacts in any category, likely reflecting a smaller population at Kiwulan at this time, making spatial patterns and hotspots difficult to discern with confidence. This testing reveals clustering of ornaments during the European period is highly non-random, potentially indicating different degrees of access to foreign ornaments or a concentration of power to control the distribution of ornaments at Kiwulan during this period.

@@

How might these results fit into a bigger picture of social change at periphery of colonial systems? We may get some insight into the general pathways that led to social inequality in northeastern Taiwan by considering how people have achieved and maintained power in a wide variety of societies [@Drennan2010; @Feinman2000; @Ames2010; @Bowles2010]. The corporate/network model proposed by @Feinman2000 expands traditional hierarchical complexity to provide a comparative basis for distinct strategies for power. In the network mode, inequality develops when individuals accumulate wealth through their individual networks and people use their wealth to attract factions, control resources, and monopolize trade networks. In contrast, the corporate mode stresses shared power across different groups and sectors, integrative ceremonies and rituals, and large cooperative labour tasks [@Feinman2000; @Siegel1999]. We may be able to interprete Yilan social organization as moving from corporate mode before the European arrival, to a network mode during European presence. The small number of ornaments, and less concentrated distribution during the pre-European period appears consistent with shared power and wealth of the corporate mode. The long-distance trade network introduced by Europeans resulted in the appearance of a network mode, and the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting trade goods [@Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

@@

We are still far from understanding the full variety of colonial impacts on peripheral indigenous communities. The origin of ornaments and how their sources changed over time would give more information to construct a clear picture of complex trade networks during this periods. By focusing on the distribution patterns in a settlement site, the Kiwulan ornaments suggest that foreign ornaments can be a proxy to detect indirect colonial influence on local indigenous populations. Ornaments give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system and use them in their daily lives to display or intensify status differences. Future work could extend this approach to sourcing the ornaments directly with geochemical methods, and studies of other trade goods such as ceramics. We have introduced here the corporate/network model for understanding the dynamics of social inequality at Kiwulan, and future tests of this should include analysis of pottery production and standardization, and mortuary practices.
